---
title: "DATATHON Final Project: Impact of historical greenhouse emissions on fish response: United States, a case study"
author: "Lesly"
date: "2024-12-17"
output: html_document
---

# Introduction

-   **Objective**: Analyze the relationship between historical emissions of CO₂, CH₄, and N₂O and fish response to climate change using publicly available datasets.
-   **Research Question**: How do historical greenhouse gas emissions (standardized to CO₂ equivalents) influence fish responses in the United States?

# Methodology

In order to conduct the analysis, the following R libraries were used:

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(rstatix)
library(knitr)
library(gganimate)
```

**Data sources**: National contributions to climate change due to historical emissions of carbon dioxide, methane, and nitrous oxide from 1830 to 2023 were obtained from (Jones et al., 2023), in an Excel document.

```{r}
emissions_data <- read_excel("USA Emissions.xlsx")
head(emissions_data)
```

Fish and climate change data were obtained from the FiCli website, which is a site that provides a comprehensive database of peer-reviewed literature available on how climate change has impacted inland fishes worldwide.The only filter used were country "United States". The document was download in a csv format.

```{r}
fish_data <- read_csv("database (3).csv")
head(fish_data)
colnames(fish_data)
```

**Standardization of greenhouse gas emissions**: To analyze emissions uniformly, all gases were converted to CO₂ equivalents (Tg/year) using global warming potentials (GWPs) (Pittelkow et al., 2013):

CH₄ (Methane): GWP = 25, N₂O (Nitrous Oxide): GWP = 298, CO₂: Directly converted from Pg to Tg by multiplying by 1,000.

All gases were added in order to have a total in CO₂ equivalents by year. 
```{r}
emissions_data <- emissions_data %>%
  mutate(
    CH4_in_CO2eq = if_else(Gas == "CH[4]", Data * 25, 0),
    N2O_in_CO2eq = if_else(Gas == "N[2]*O", Data * 298, 0),
    CO2_in_CO2eq = if_else(Gas == "CO[2]", Data * 1000, 0),
    Total_CO2eq = CH4_in_CO2eq + N2O_in_CO2eq + CO2_in_CO2eq
  ) %>%
  group_by(Year) %>%
  summarize(Total_CO2eq = sum(Total_CO2eq, na.rm = TRUE))
head(emissions_data)
summary(emissions_data)
```


**Data integration**: The emissions dataset was merged with the fish response dataset based on the "Year" variable. As fish response data ranged from 1964–2020 with gaps, the merged dataset only included overlapping years with available data from both sources.
```{r}
merged_data <- fish_data %>%
  mutate(Fish_Response = as.factor(Fish_Response)) %>%
  inner_join(emissions_data, by = "Year")
```

**Explanatory Data Analysis**

**I) Emissions Time Series**: Emissions data were visualized using animated time-series plot to highlight trends in CO₂-equivalent emissions over time.
```{r}
ggplot(emissions_data, aes(x = Year, y = Total_CO2eq)) +
  geom_line(color = "blue") +
  labs(
    title = "Animated Emissions Over Time",
    x = "Year",
    y = "Total CO2eq (Tg/year)"
  ) +
  transition_reveal(Year) +
  theme_minimal()
```

**II) Fish response**
Fish response categories were summarized using bar plots to identify the distribution of responses.
```{r}
fish_data %>% count(Fish_Response)
```


**Statistical Analysis**: A logistic regression model was employed to examine the relationship between total CO₂-equivalent emissions (predictor variable) and fish response categories (dependent variable).
```{r}
model <- glm(Fish_Response ~ Total_CO2eq, 
             data = merged_data, 
             family = "binomial")
```




# Results

-   **Emission trends**: The time series of emissions shows trends in CO₂-equivalent emissions over time. 

```{r}
ggplot(emissions_data, aes(x = Year)) +
  geom_line(aes(y = Total_CO2eq, color = "Total CO2eq (Tg/year)")) +
  labs(
    title = "Historical USA Emissions in CO2 Equivalents",
    x = "Year",
    y = "Emissions (Tg CO2eq/year)",
    color = "Greenhouse gas"
  ) +
  theme_minimal()
```


-   **Fish Response**: The categorical distribution of fish responses suggests potential patterns in relation to emissions.
```{r}
ggplot(fish_data, aes(x = Fish_Response)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Fish Response Categories",
    x = "Fish Response Category",
    y = "Count"
  ) +
  theme_minimal()
```

-   **Statistical Model**: The logistic regression model indicates significant relationships between total CO₂-equivalent emissions and fish response (details in summary table).
```{r}
summary(model)
```


### Visualize the relationship

```{r}
ggplot(merged_data, aes(x = Total_CO2eq, y = Fish_Response)) +
  geom_jitter(width = 0.1, height = 0.1, alpha = 0.6) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), color = "blue") +
  labs(
    title = "Relationship Between Total CO2eq and Fish Response",
    x = "Total CO2eq (Tg/year)",
    y = "Fish Response (log-odds)"
  ) +
  theme_minimal()
```


# Discussion

-   **Findings**:
    -   Total CO₂-equivalent emissions are significantly correlated with changes in fish responses.
    -   Increasing emissions are associated with more negative fish response categories.
-   **Limitations**:
    -   Potential biases in the fish response dataset.
    -   Lack of control variables (e.g., temperature).

# Conclusions

-   This study highlights the impact of historical greenhouse gas emissions (in CO₂ equivalents) on fish response.

-   Future work should include additional ecological variables for a more comprehensive analysis.

-   **References**:

1.  Jones, M. W., Peters, G. P., Gasser, T., Andrew, R. M., Schwingshackl, C., Gütschow, J., Houghton, R. A., Friedlingstein, P., Pongratz, J., & Quéré, C. L. (2023). National contributions to climate change due to historical emissions of carbon dioxide, methane, and nitrous oxide since 1850. Scientific Data, 10(1). <https://doi.org/10.1038/s41597-023-02041-1>

2.  Krabbenhoft, T. J., Myers, B. J. E., Wong, J. P., Chu, C., Tingley, R. W., Falke, J. A., Kwak, T. J., Paukert, C. P., & Lynch, A. J. (2020). FiCli, the Fish and Climate Change Database, informs climate adaptation and management for freshwater fishes. Scientific Data, 7(1). <https://doi.org/10.1038/s41597-020-0465-z>

3.  Fish and Climate Change Database (FICLI). (n.d.). <https://rconnect.usgs.gov/ficli/>

4.  Pittelkow, C. M., Adviento-Borbe, M. A., Hill, J. E., Six, J., Van Kessel, C., & Linquist, B. A. (2013). Yield-scaled global warming potential of annual nitrous oxide and methane emissions from continuously flooded rice in response to nitrogen input. Agriculture Ecosystems & Environment, 177, 10–20. <https://doi.org/10.1016/j.agee.2013.05.011>

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
