---
title: "DATATHON Final Project: Impact of historical greenhouse emissions on fish response: A case study in United States"
author: "Lesly"
date: "2024-12-17"
output: html_document
---

# Introduction

-   **Objective**: Analyze the relationship between historical emissions of CO₂, CH₄, and N₂O and fish response to climate change using publicly available datasets.
-   **Research Question**: How do historical greenhouse gas emissions (standardized to CO₂ equivalents) influence fish responses in the United States?


# Methodology

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(ggplot2)
library(dplyr)
library(rstatix)
library(knitr)
library(gganimate)
```


## Load data
### Emissions from 1830 to 2023
```{r}
emissions_data <- read_excel("USA Emissions.xlsx")
head(emissions_data)
```


## Fish data
```{r}
fish_data <- read_csv("database (3).csv")
head(fish_data)
colnames(fish_data)
```





-   **References**:
1. Jones, M. W., Peters, G. P., Gasser, T., Andrew, R. M., Schwingshackl, C., Gütschow, J., Houghton, R. A., Friedlingstein, P., Pongratz, J., & Quéré, C. L. (2023). National contributions to climate change due to historical emissions of carbon dioxide, methane, and nitrous oxide since 1850. Scientific Data, 10(1). https://doi.org/10.1038/s41597-023-02041-1

2. Krabbenhoft, T. J., Myers, B. J. E., Wong, J. P., Chu, C., Tingley, R. W., Falke, J. A., Kwak, T. J., Paukert, C. P., & Lynch, A. J. (2020). FiCli, the Fish and Climate Change Database, informs climate adaptation and management for freshwater fishes. Scientific Data, 7(1). https://doi.org/10.1038/s41597-020-0465-z

    
    
    
```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
